# Introdución {#intro}

Trabajar con la interfaz de gráfica de R es un poco incómodo (poniendolo 
suavemente). Por lo que se han desarrollado varios entornos de desarrollo  
integrado (IDE por sus siglas en inglés) que facilitan su uso. Una de las más
trabajadas y pulidas es RStudio.

## Introducción a RStudio

RStudio es una interfaz de desarrollo integrada, gratis y de código abierto.
Posee un editor de código, funciona en múltiples plataformas (incluyendo
servidores) y tiene muchas ventajas como integración con control de versión y
manejo de proyectos.


**Disposición Básica**

Cuando abres RStudio por primera vez verás tres paneles:

  +  La consola interactiva de R (a la izquierda)
  +  El Entorno/Historial, en inglés Environment/History (en pestañas en la parte superior derecha)
  +  Archivos/Gráficos/Paquetes/Ayuda/Visualizador, también en inglés Files/Plots/Packages/Help/Viewer (en la parte inferior derecha)
  
Una vez que abre un archivo, como por ejemplo un script de R, un panel de edición
se abre en la parte superior izquierda.

## Flujo de trabajo con RStudio
Hay dos maneras básicas de trabajar con RStudio.

1. Probar y jugar dentro de la consola interactiva y luego copiar el código a un
archivo .R para ejecutar más tarde.
   *  Funciona bien cuando se hacen pequeñas pruebas y recién se comienza.
   *  Pero rápidamente se vuelve laborioso.
2. Empezar a escribir un archivo .R y luego usar el comando / atajo de RStudio para
ejecutar la línea actual o las líneas seleccionadas en la consola.
   * Es la mejor forma de empezar; así todo el código queda guardado para más
   tarde
   * Podrás ejecutar el archivo desde RStudio o usando la función de R
   `source()`.
   
   
```{block, running-code-seg, type = 'rmdtip'}
## Tip: Ejecutando partes de tu código

RStudio tiene una gran flexibilidad al ejecutar código dentro de la ventana
de edición. Hay botones, menues y atajos de teclado. Para ejectuar la línea
actual podés 1. hacer click en el botón `Run` arriba del panel de edición, o
2. Seleccionar "Run Lines" desde el menú "Code", o 3. apretar Ctrl+Enter en 
Windows o Linux o Command+Enter en OS X (Este atajo también puede verse 
dejando el puntero del mouse sobre el botón `Run`). Para ejecutar un bloque de
código, seleccionálo y después solo presiona `Run`. Si has modificado una parte
del bloque que has ejecutado reción no hay necesidad de reseleccionarlo y 
darle `Run`, podés usar el botón a continuación, `Re-run the previous region`.
Esto va a ejecutar el bloque de código previo incluyendo las modificaciones que
has hecho.
```

## Introducción a R

Una gran parte del tiempo en R se pasa en la consola interactiva. Ahí es donde
se corre el código y puede ser un entorno útil para probar ideas antes de 
escribirlas en un script de R. La consola de RStudio es la misma que si ejecutasen
`R` en la linea de comando o la interfaz gráfica de R (R-GUI). 

Cada vez que abran la consola van a ver un montón de información y abajo aparece
un signo mayor ">" seguido de un cursor parpadeante. Opera en la forma como 
cualquier otra línea de comando. REPL por sus siglas en inglés: Read, evaluate, 
print, loop. Leer, evaluar, imprimir, buclear. Uno ingresa órdenes o comandos,
R trata de ejecutarlos y luego imprime el resultado. Aunque a veces algunas
funciones no imprimen nada.


## R como calculadora


Lo más simple que se puede hacer con R es aritmética:


```{r c1-aritmetica-1, results='asis'}
1 + 100
```

Y R va a impirmir la respuesta, precedida por un "[1]". No se preocupen por esto
ahora, lo vamos a ver más adelante. Por ahora piensen que está indicando la
salida.

Si introducen un comando de forma incompleta, R va esperar que lo terminen

```{r c1-entrada-incompleta-1, eval=FALSE}
1 +
```

```{r c1-entrada-incompleta-2, eval=FALSE}
+
```

Cada vez que apriten Enter y la consola muestre un "+" en vez de ">", es que
está esperando que terminen el comando. Si quieren cancelar el comando solo
hay que presionar "Esc" y la consola volverá a mostrar el prompt ">".


```{block type='callout'}
## Tip: Cancelando comandos

Si estás usando R desde la linea de comando en vez de dentro de RStudio,
hay que presionar `Ctrl+C` en vez de  `Esc` para cancelar el comando. Esto
también se aplica a usuarios de Mac también.

Cancelar un comando no es útil solamente para anular comandos incompletos:
también podés cancelar una computación que R está ejecutando (por ejemplo, 
si está tomando demasiado tiempo), o para eliminar el código que estás
escribiendo actualmente.
```


Cuando se usa R como calculadora, el orden las operaciones es el mismo que 
aprendieron en la escuela.

De mayor a menor precedencia:

  + Paréntesis: `(`, `)`
  + Exponentes: `^` o `**`
  +
When using R as a calculator, the order of operations is the same as you
would have learned back in school.

From highest to lowest precedence:

 * Parentheses: `(`, `)`
 * Exponents: `^` or `**`
 * Divide: `/`
 * Multiply: `*`
 * Add: `+`
 * Subtract: `-`


~~~
3 + 5 * 2
~~~
{: .r}



~~~
[1] 13
~~~
{: .output}

Use parentheses to group operations in order to force the order of
evaluation if it differs from the default, or to make clear what you
intend.


~~~
(3 + 5) * 2
~~~
{: .r}



~~~
[1] 16
~~~
{: .output}

This can get unwieldy when not needed, but  clarifies your intentions.
Remember that others may later read your code.


~~~
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if you remember the rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
~~~
{: .r}


The text after each line of code is called a
"comment". Anything that follows after the hash (or octothorpe) symbol
`#` is ignored by R when it executes code.

Really small or large numbers get a scientific notation:


~~~
2/10000
~~~
{: .r}



~~~
[1] 2e-04
~~~
{: .output}

Which is shorthand for "multiplied by `10^XX`". So `2e-4`
is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:


~~~
5e3  # Note the lack of minus here
~~~
{: .r}



~~~
[1] 5000
~~~
{: .output}

## Mathematical functions

R has many built in mathematical functions. To call a function,
we simply type its name, followed by  open and closing parentheses.
Anything we type inside the parentheses is called the function's
arguments:


~~~
sin(1)  # trigonometry functions
~~~
{: .r}



~~~
[1] 0.841471
~~~
{: .output}


~~~
log(1)  # natural logarithm
~~~
{: .r}



~~~
[1] 0
~~~
{: .output}


~~~
log10(10) # base-10 logarithm
~~~
{: .r}



~~~
[1] 1
~~~
{: .output}


~~~
exp(0.5) # e^(1/2)
~~~
{: .r}



~~~
[1] 1.648721
~~~
{: .output}

Don't worry about trying to remember every function in R. You
can simply look them up on Google, or if you can remember the
start of the function's name, use the tab completion in RStudio.

This is one advantage that RStudio has over R on its own, it
has auto-completion abilities that allow you to more easily
look up functions, their arguments, and the values that they
take.

Typing a `?` before the name of a command will open the help page
for that command. As well as providing a detailed description of
the command and how it works, scrolling to the bottom of the
help page will usually show a collection of code examples which
illustrate command usage. We'll go through an example later.

## Comparing things

We can also do comparison in R:


~~~
1 == 1  # equality (note two equals signs, read as "is equal to")
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}


~~~
1 != 2  # inequality (read as "is not equal to")
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}


~~~
1 <  2  # less than
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}


~~~
1 <= 1  # less than or equal to
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}


~~~
1 > 0  # greater than
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}


~~~
1 >= -9 # greater than or equal to
~~~
{: .r}



~~~
[1] TRUE
~~~
{: .output}

> ## Tip: Comparing Numbers
>
> A word of warning about comparing numbers: you should
> never use `==` to compare two numbers unless they are
> integers (a data type which can specifically represent
> only whole numbers).
>
> Computers may only represent decimal numbers with a
> certain degree of precision, so two numbers which look
> the same when printed out by R, may actually have
> different underlying representations and therefore be
> different by a small margin of error (called Machine
> numeric tolerance).
>
> Instead you should use the `all.equal` function.
>
> Further reading: [http://floating-point-gui.de/](http://floating-point-gui.de/)
>
{: .callout}

## Variables and assignment

We can store values in variables using the assignment operator `<-`, like this:


~~~
x <- 1/40
~~~
{: .r}

Notice that assignment does not print a value. Instead, we stored it for later
in something called a **variable**. `x` now contains the **value** `0.025`:


~~~
x
~~~
{: .r}



~~~
[1] 0.025
~~~
{: .output}

More precisely, the stored value is a *decimal approximation* of
this fraction called a [floating point number](http://en.wikipedia.org/wiki/Floating_point).

Look for the `Environment` tab in one of the panes of RStudio, and you will see that `x` and its value
have appeared. Our variable `x` can be used in place of a number in any calculation that expects a number:


~~~
log(x)
~~~
{: .r}



~~~
[1] -3.688879
~~~
{: .output}

Notice also that variables can be reassigned:


~~~
x <- 100
~~~
{: .r}

`x` used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:


~~~
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
~~~
{: .r}

The right hand side of the assignment can be any valid R expression.
The right hand side is *fully evaluated* before the assignment occurs.

Variable names can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces at all. Different people use
different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but **be consistent**.

It is also possible to use the `=` operator for assignment:


~~~
x = 1/40
~~~
{: .r}

But this is much less common among R users.  The most important thing is to
**be consistent** with the operator you use. There are occasionally places
where it is less confusing to use `<-` than `=`, and it is the most common
symbol used in the community. So the recommendation is to use `<-`.

## Vectorization

One final thing to be aware of is that R is *vectorized*, meaning that
variables and functions can have vectors as values. For example


~~~
1:5
~~~
{: .r}



~~~
[1] 1 2 3 4 5
~~~
{: .output}



~~~
2^(1:5)
~~~
{: .r}



~~~
[1]  2  4  8 16 32
~~~
{: .output}



~~~
x <- 1:5
2^x
~~~
{: .r}



~~~
[1]  2  4  8 16 32
~~~
{: .output}

This is incredibly powerful; we will discuss this further in an
upcoming lesson.


## Managing your environment

There are a few useful commands you can use to interact with the R session.

`ls` will list all of the variables and functions stored in the global environment
(your working R session):


~~~
ls()
~~~
{: .r}



~~~
[1] "x" "y"
~~~
{: .output}

> ## Tip: hidden objects
>
> Like in the shell, `ls` will hide any variables or functions starting
> with a "." by default. To list all objects, type `ls(all.names=TRUE)`
> instead
>
{: .callout}

Note here that we didn't given any arguments to `ls`, but we still
needed to give the parentheses to tell R to call the function.

If we type `ls` by itself, R will print out the source code for that function!


~~~
ls
~~~
{: .r}



~~~
function (name, pos = -1L, envir = as.environment(pos), all.names = FALSE, 
    pattern, sorted = TRUE) 
{
    if (!missing(name)) {
        pos <- tryCatch(name, error = function(e) e)
        if (inherits(pos, "error")) {
            name <- substitute(name)
            if (!is.character(name)) 
                name <- deparse(name)
            warning(gettextf("%s converted to character string", 
                sQuote(name)), domain = NA)
            pos <- name
        }
    }
    all.names <- .Internal(ls(envir, all.names, sorted))
    if (!missing(pattern)) {
        if ((ll <- length(grep("[", pattern, fixed = TRUE))) && 
            ll != length(grep("]", pattern, fixed = TRUE))) {
            if (pattern == "[") {
                pattern <- "\\["
                warning("replaced regular expression pattern '[' by  '\\\\['")
            }
            else if (length(grep("[^\\\\]\\[<-", pattern))) {
                pattern <- sub("\\[<-", "\\\\\\[<-", pattern)
                warning("replaced '[<-' by '\\\\[<-' in regular expression pattern")
            }
        }
        grep(pattern, all.names, value = TRUE)
    }
    else all.names
}
<bytecode: 0x7feecdab3f88>
<environment: namespace:base>
~~~
{: .output}

You can use `rm` to delete objects you no longer need:


~~~
rm(x)
~~~
{: .r}

If you have lots of things in your environment and want to delete all of them,
you can pass the results of `ls` to the `rm` function:


~~~
rm(list = ls())
~~~
{: .r}

In this case we've combined the two. Like the order of operations, anything
inside the innermost parentheses is evaluated first, and so on.

In this case we've specified that the results of `ls` should be used for the
`list` argument in `rm`. When assigning values to arguments by name, you *must*
use the `=` operator!!

If instead we use `<-`, there will be unintended side effects, or you may get an error message:


~~~
rm(list <- ls())
~~~
{: .r}



~~~
Error in rm(list <- ls()): ... must contain names or character strings
~~~
{: .error}

> ## Tip: Warnings vs. Errors
>
> Pay attention when R does something unexpected! Errors, like above,
> are thrown when R cannot proceed with a calculation. Warnings on the
> other hand usually mean that the function has run, but it probably
> hasn't worked as expected.
>
> In both cases, the message that R prints out usually give you clues
> how to fix a problem.
>
{: .callout}

## R Packages

It is possible to add functions to R by writing a package, or by
obtaining a package written by someone else. As of this writing, there
are over 7,000 packages available on CRAN (the comprehensive R archive
network). R and RStudio have functionality for managing packages:

* You can see what packages are installed by typing
  `installed.packages()`
* You can install packages by typing `install.packages("packagename")`,
  where `packagename` is the package name, in quotes.
* You can update installed packages by typing `update.packages()`
* You can remove a package with `remove.packages("packagename")`
* You can make a package available for use with `library(packagename)`

> ## Challenge 1
>
> Which of the following are valid R variable names?
> 
> ~~~
> min_height
> max.height
> _age
> .mass
> MaxLength
> min-length
> 2widths
> celsius2kelvin
> ~~~
> {: .r}
>
> > ## Solution to challenge 1
> >
> > The following can be used as R variables:
> > 
> > ~~~
> > min_height
> > max.height
> > MaxLength
> > celsius2kelvin
> > ~~~
> > {: .r}
> >
> > The following creates a hidden variable:
> > 
> > ~~~
> > .mass
> > ~~~
> > {: .r}
> >
> > The following will not be able to be used to create a variable
> > 
> > ~~~
> > _age
> > min-length
> > 2widths
> > ~~~
> > {: .r}
> {: .solution}
{: .challenge}

> ## Challenge 2
>
> What will be the value of each  variable  after each
> statement in the following program?
>
> 
> ~~~
> mass <- 47.5
> age <- 122
> mass <- mass * 2.3
> age <- age - 20
> ~~~
> {: .r}
>
> > ## Solution to challenge 2
> >
> > 
> > ~~~
> > mass <- 47.5
> > ~~~
> > {: .r}
> > This will give a value of 47.5 for the variable mass
> >
> > 
> > ~~~
> > age <- 122
> > ~~~
> > {: .r}
> > This will give a value of 122 for the variable age
> >
> > 
> > ~~~
> > mass <- mass * 2.3
> > ~~~
> > {: .r}
> > This will multiply the existing value of 47.5 by 2.3 to give a new value of
> > 109.25 to the variable mass.
> >
> > 
> > ~~~
> > age <- age - 20
> > ~~~
> > {: .r}
> > This will subtract 20 from the existing value of 122 to give a new value
> > of 102 to the variable age.
> {: .solution}
{: .challenge}


> ## Challenge 3
>
> Run the code from the previous challenge, and write a command to
> compare mass to age. Is mass larger than age?
>
> > ## Solution to challenge 3
> >
> > One way of answering this question in R is to use the `>` to set up the following:
> > 
> > ~~~
> > mass > age
> > ~~~
> > {: .r}
> > 
> > 
> > 
> > ~~~
> > [1] TRUE
> > ~~~
> > {: .output}
> > This should yield a boolean value of TRUE since 109.25 is greater than 102.
> {: .solution}
{: .challenge}


> ## Challenge 4
>
> Clean up your working environment by deleting the mass and age
> variables.
>
> > ## Solution to challenge 4
> >
> > We can use the `rm` command to accomplish this task
> > 
> > ~~~
> > rm(age, mass)
> > ~~~
> > {: .r}
> {: .solution}
{: .challenge}

> ## Challenge 5
>
> Install the following packages: `ggplot2`, `plyr`, `gapminder`
>
> > ## Solution to challenge 5
> >
> > We can use the `install.packages()` command to install the required packages.
> > We can also install them in one go like so:
> > 
> > ~~~
> > install.packages(c("ggplot2", "plyr", "gapminder"))
> > ~~~
> > {: .r}
> {: .solution}
{: .challenge}
